Converging High-Throughput and High-Performance Computing: A Case Study

### 2017-04-02 introduction.tex
minor refinement of introduction.
  remove abstract to give to Jack et al.

The Large Hadron Collider (LHC) was created to explore the fundamental
properties of matter for the next decades.  Since LHC start-up in 2009,
multiple experiments  at LHC have collected and distributed hundreds of
petabytes of data worldwide to hundreds of computer centers. Thousands of
physicists analyze petascale data volumes daily. The detection of the Higgs
Boson speaks to the success of the detector and experiment design, as well as
the sophistication of computing systems devised to analyze the data.

Historically the computing systems consisted of the federation of a hundreds
to thousands of distributed resources \textemdash{} ranging in scale from
small to mid-size resource~\cite{foster2003grid}. Although the workloads to be
executed are independent, the management of the distribution of extremely-
large workloads across many heterogeneous resources to ensure the effective
utilization of resources and efficient execution of workloads presents non-
trivial challenges.

Many software solutions have been developed in response to these challenges.
One of the LHC experiments, the CMS experiment devised a solution based around
the HTCondor~\cite{XX} software ecosystem. The ATLAS~\cite{Aad:2008}
experiment, utilizes the Production and Distributed Analysis (PanDA) workload
management system~\cite{Maeno2011} (WMS) for distributed data processing and
analysis. The CMS and ATLAS experiments represent arguably the largest
production grade distributed computing solutions and have symbolized the
paradigm of {\it high-throughput computing} viz., the effective execution of
many independent workloads.

As the LHC prepares for the high-luminosity era (Run 3 in $~\approx$ 2022), it
is anticipated that the data volumes that will need analyzing will increase by
10X compared to the current phase (Run 2).   The data will be larger in volume
but will also require heterogeneous computational processes. In spite of the
impressive scale of the ATLAS distributed computing system, demand for
computing systems will significantly outstrip supply (availability).

@@

This paper is a case study of how the ATLAS experiment has "broken free" of
the traditional computational approach of high-throughput computing on
distributed resources to embrace new platforms, in particular high-performance
computers. Specifically, we discuss the experience of integrating the PanDA
workload management system with Titan ~\textemdash{} a DOE leadership
computing facility and scaled to analyze up to \jhanote{XX\%} of the ATLAS
experiment workload. We present the design of PanDA and show how localized
substitution to use well established pilot-concepts allow enhanced support for
heterogeneous workloads (such as molecular dynamics) and advanced execution
modes.

This state-of-practice paper provides multiple contributions.  (i) Documents
the many design and operational considerations that have been  taken to
support the sustained, scalable and production usage of Titan for historically
high-throughput workloads, (ii) Extensions to PanDA to support non-traditional
heterogeneous workloads and execution modes,  and (iii) As the community looks
forward to designing the next generation of online analytical platforms
~\cite{foap} the lessons learnt from our project provide some guidance for how
current and future experimental and observational systems can be integrated
with supercomputers in production.

>>>>

The Large Hadron Collider (LHC) was created to explore the fundamental
properties of matter for the next decades.   Multiple experiments at LHC have
collected and distributed hundreds of petabytes of data worldwide to hundreds
of computer centers. Thousands of physicists analyze petascale data volumes
daily. The detection of the Higgs Boson in 2013 speaks to the success of the
detector and experiment design, as well as the sophistication of computing
systems devised to analyze the data.

Historically the computing systems used by LHC experiments consisted of the
federation of a hundreds to thousands of distributed resources \textemdash{}
ranging in scale from small to mid-size resource~\cite{foster2003grid}.
Although the workloads to be executed are comprised of tasks that are
independent of each other, the management of the distribution of 
workloads across many heterogeneous resources to ensure the effective
utilization of resources and efficient execution of workloads presents non-
trivial challenges.

Many software solutions have been developed in response to these challenges.
One of the LHC experiments, the CMS experiment devised a solution based around
the HTCondor~\cite{XX} software ecosystem. The ATLAS~\cite{Aad:2008}
experiment, utilizes the Production and Distributed Analysis (PanDA) workload
management system~\cite{Maeno2011} (WMS) for distributed data processing and
analysis. The CMS and ATLAS experiments represent arguably the largest
production grade distributed computing solutions and have symbolized the
paradigm of {\it high-throughput computing} viz., the effective execution of
many independent tasks.

As the LHC prepares for the high-luminosity era (Run 4) and Run 3 in
$~\approx$ 2022, it is anticipated that the data volumes that will need
analyzing will increase by factors of 10-100  compared to the current phase
(Run 2). The data will be larger in volume but will also require more
sophisticated computational processing. In spite of the impressive scale of
the ATLAS distributed computing system, demand for computing systems will
significantly outstrip supply (availability).

@@

This paper is a case study of how the ATLAS experiment has "broken free" of
the traditional computational approach of high-throughput computing on
distributed resources to embrace new platforms, in particular high-performance
computers. Specifically, we discuss the experience of integrating the PanDA
workload management system with Titan ~\textemdash{} a DOE leadership
computing facility. Consequently, Titan now analyzes up to 5\% of workload  of
the ATLAS experiment. The case study presents the design and integration of
PanDA to reach sustained production scales (ca. 100 M core-hours a years). It
also discusses the investigation of a task execution runtime system on Titan,
that is based on  the pilot-abstractions and which allows advanced execution
modes of the ATLAS workload as well as the enhanced support for heterogeneous
workloads (such as molecular dynamics).

This state-of-practice paper provides multiple contributions.  It (i)
documents the many design and operational considerations that have been  taken
to support the sustained, scalable and production usage of Titan for
historically high-throughput workloads, (ii) Extensions to PanDA to support
non-traditional heterogeneous workloads and execution modes,  and (iii) As the
community looks forward to designing the next generation of online analytical
platforms~\cite{foap}, this project provides some early lessons and guidance
for how current and future experimental and observational systems can be
integrated with production supercomputers in a general and extensible manner.
